FITS format for planetary surfaces: definitions, applications and best practices.

### 2018-02-09 ess_marmo.tex
Update on Overleaf.

% (A title should be specific, informative, and brief. Use
% abbreviations only if they are defined in the abstract. Titles that
% start with general terms then specific results are optimized in
% searches)

% Example: \title{This is a test title}

@@

%%% Suggested section heads:
% \section{Introduction}
% 
% The main text should start with an introduction. Except for short
% manuscripts (such as comments and replies), the text should be divided
% into sections, each with its own heading.

% Headings should be sentence fragments and do not begin with a
% lowercase letter or number. Examples of good headings are:

% \section{Materials and Methods}
% Here is text on Materials and Methods.

% \subsection{A descriptive heading about methods}
% More about Methods.
% 
% \section{Data} (Or section title might be a descriptive heading about data)
% 
% \section{Results} (Or section title might be a descriptive heading about the
% results)
% 
% \section{Conclusions}

@@

\section{FITS format in Planetary Surface investigation}
\label{sec:fitspss}

\subsection{FITS format}
FITS is an open digital standard, defined for data acquisition and archiving
in astronomical observatories in the late 70's, and it is used for
spatial telescope data.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard implemented formats in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO).
Finally, FITS format is compatible with the Planetary Data System (PDS)
archiving specifications and supported by a large number of open libraries
and software tools.

@@

\subsection{Planetary Surface proposed convention}
\label{ssec:geofits}
When willing to use FITS format for planetary surfaces the issue of rigorous
geographical representation must be addressed.
For their mapping and analysis needs planetary researchers commonly use
Geographical Information Systems (GIS).
In general, GIS applications excel in data interoperability even
though some have historically been anchored to Earth's spatial description.
Several of them are based on the Geospatial Data Abstraction
Library\footnote{http://www.gdal.org} (GDAL), released by the Open Source
Geospatial Foundation.
GDAL is essentially a format translation library written in C++ for geospatial raster
and vector data and, fortunately, it also offer basic support for FITS format.

@@

\begin{table}
\label{table:proj}
\caption{Projection names and codes already available in FITS suitable for
planetary applications: their correspondent in PDS4 vocabulary are
provided.}
\centering
\begin{tabular}{l l l}
\hline
FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
\hline
Zenithal Equidistant & ARC & Azimuthal Equidistant \\
Zenithal Perspective & AZP & General Vertical Near-sided Projection \\
Plate Carre & CAR & Equirectangular, Miller Cylindrical \\
Equidistant Conic & COD & Equidistant Conic \\
Conic Equal-Area & COE & Albers Conical Equal Area \\ 
Conic Ortomorphic & COO & Lambert Conformal Conic\\
%Cylindrical Equal-Area & CEA & \textit{there is no CEA option in PDS4 neither in ISIS3}\\
Mercator & MER & Mercator, Oblique Mercator, Transverse Mercator \\
Polyconic & PCO & Polyconic \\
Sanson-Flamsteed & SFL & Sinusoidal \\
Ortographic & SIN & Orthographic \\
Stereographic & STG & Stereographic, Polar Stereographic \\
Gnomonic & TAN & Gnomonic \\
Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
\hline
\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}. We cannot find correspondences in the WCS scheme for PDS4 Robinson,} \\
\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator.}
\end{tabular}
\end{table}

@@

\subsubsection{Correspondence between reserved FITS keywords and PDS4 standard reference}
\label{sssec:stdfitspds4}
The correspondence between mandatory and reserved FITS keywords (as defined in
the last version of the FITS standard document\footnote{https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf})
and PDS4 attributes, is described here.
\begin{itemize}
\item{\texttt{BITPIX}.
It specifies the number of bits that represent a data value in
the associated data array.
PDS4 uses the attribute \texttt{data\_type}: correspondent 
values depend on the associated object class.}

@@

\end{itemize}

@@

\subsubsection{Visualization of global and projected maps}
High level products are meant to be used for visual inspection or GIS analysis.
They are distributed as projected rasters, often in GeoTIFF\footnote{http://geotiff.osgeo.org/}
format, as GeoTIFF make them GIS ready.
Once new keywords introduced in FITS and WCS definitions adapted to planetary surface, FITS
and GeoTIFF are completely interchangeable.
As an example we have downloaded some global maps from USGS
Astropedia\footnote{https://astrogeology.usgs.gov/search} and converted them in FITS.
We used ds9 to display the FITS image and were able to visualize coordinates in meters
as in QGIS, but also in planetary latitude and longitude thanks to the alternate coordinate
description (see figure \ref{fig:ds9qgis}, left side).
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
\caption{left: Mercury South Pole FITS image displayed in ds9;
right: GeoTIFF raster superposed to the virtual
description of the FITS raster.
Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied Physics
Laboratory, Carnegie Science}
\label{fig:ds9qgis}
\end{figure}

@@

\subsubsection{OMEGA spectra and geometry data}
As stressed above, having planetary data distributed in FITS at any processing level is
a way to avoid multiple conversions in the processing chain and simplify reprocessing.
In planetary science a first spatial calibration data often consists in providing coordinates
on specific points on the detector (the center of the pixel, one or more pixel corners, etc.).
To support such look-up table coordinate representation the \texttt{TAB} algorithm
has been defined in \citet{wcsspectral}.
In the \texttt{TAB} algorithm, coordinates are listed in a coordinate array, an indexing vector
can be used to address coordinate array elements.
Coordinates can then be sampled more or less coarsely depending on the behavior of the spatial
reconstruction.
Also, when the field is only partially covered by the planetary surface, coordinate array
dimensions can be significantly reduced.

As an example of non projected planetary data we have converted in FITS format data
acquired by the spectroimager OMEGA on-board of Mars Express.
OMEGA unprojected and uncalibrated data are distributed by ESA-PSA and PDS archives.
Their structure is described in the OMEGA Experiment Archive 
Interface Control Document\footnote{ftp://psa.esac.esa.int/pub/mirror/MARS-EXPRESS/OMEGA/MEX-M-OMEGA-2-EDR-FLIGHT-V1.0/DOCUMENT/EAICD\_OMEGA.PDF}.
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo4}\hspace{60pt}\includegraphics[height=17pc]{ess_marmo3}}
\caption{left: OMEGA data structure (from OMEGA EAICD); right: FITS Multi Digital Object
conversion}
\label{fig:omegascheme}
\end{figure}

TAB projection is implemented in the Calabretta WCSLIB\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/},
available in all major linux distribution, and consequently in all

\subsubsection{Akatsuki imaging data}

\subsubsection{HiRISE mosaics}

\section{Inter-operable software developments}
\label{sec:softdev}
VO-QGIS-plugin\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}

@@

\section{Discussion and Perspectives}
\label{sec:disc}
FITS discussion in astronomy\citep{THOMAS2015133} 
let's astronomy learn from planetary sciences too: accent must be put on interoperability
.... blabla
Planetary surface is lacking open and modular software for massive processing and can benefit
of astronomy experience. In addition to that WCS projection scheme is simpler than the
historical one used in planetary applications. 
Archiving flexibility and data model are today approached by PDS4 and EPN-TAP efforts.
Reducing data format conversion during the processing chain but
allowing easy conversion for final products following the needed application.

%% Enter Figures and Tables near as possible to where they are first mentioned:
%
% DO NOT USE \psfrag or \subfigure commands.
%
% Figure captions go below the figure.
% Table titles go above tables;  other caption information
%  should be placed in last line of the table, using
% \multicolumn2l{$^a$ This is a table note.}
%
%----------------
% EXAMPLE FIGURE
%
% \begin{figure}
% \end{figure}

%\begin{figure}[ht!]
% \centering
% when using pdflatex, use pdf file:
% \includegraphics[width=20pc]{figsamp.pdf}
%
% when using dvips, use .eps file:
% \includegraphics[width=20pc]{figsamp.eps}
%
% If you don't specify the file extension, 
% \includegraphics will insert the right file:
%\centerline{\includegraphics[height=1.5in]{figsamp}}
%\caption{Short caption}
%\label{figone}
%\end{figure}

%\begin{figure}[ht!]
%\centerline{\includegraphics[angle=45,height=2in]{figsamp}}
%\caption{The figure caption should begin with an overall descriptive
%statement of the figure followed by additional text. They should be
%immediately after each figure.  Figure parts are indicated with
%lower-case letters ({\bf a, b, c}\ldots).  For initial submission, please place
%both the figures and captions in the text near where they are cited.}
%\label{fig:mean_and_slope}
%\end{figure}

%\begin{table}
%\caption{Start this caption with a short description of your
%table.
%Large tables
%especially presenting rich data should be presented as separate excel
%or .cvs files, not as part of the main text.}
%== Table Here ==
%\end{table}

%
% ---------------
% EXAMPLE TABLE
%
%\begin{table}
%\caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
%\centering
%\begin{tabular}{l c}
%\hline
% Run  & Time (min)  \\
%\hline
%  $l1$  & 260   \\
%  $l2$  & 300   \\
%  $l3$  & 340   \\
%  $h1$  & 270   \\
%  $h2$  & 250   \\
%  $h3$  & 380   \\
%  $r1$  & 370   \\
%  $r2$  & 390   \\
%\hline
%\multicolumn{2}{l}{$^{a}$Table note text here.}
%\end{tabular}
%\end{table}

% See below for how to make sideways figures or tables.

% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
%
% \begin{sidewaysfigure}
% \includegraphics[width=20pc]{figsamp}
% \caption{Caption here}
% \label{newfig}
% \end{sidewaysfigure}
%
% \begin{sidewaystable}
% \caption{Caption here}
%\label{tab:signif_gap_clos}
% \begin{tabular}{ccc}
%one&two&three\\
%four&five&six
% \end{tabular}
% \end{sidewaystable}

%\clearpage

%% ------------------------------------------------------------------------ %%
%\section*{Citations}

% Please use ONLY \citet and \citep for reference citations.
% DO NOT use other cite commands (e.g., \cite, \citeyear, \nocite, \citealp, etc.).

%\subsection*{Cites made with \tt\string\citet\string{\string}}
%...as shown by \citet{Boug10}, \citet{Buiz07}, \citet{Fra10},
%\citet{Ghel00}, and \citet{Leit74}.

%\subsection*{Cites made with \tt\string\citep\string{\string}}
%...as shown by \citep{Boug10}, \citep{Buiz07}, \citep{Fra10},
%\citep{Ghel00, Leit74}.

%...has been shown \citep [e.g.,][]{Boug10,Buiz07,Fra10}.

>>>>

@@

@@

\section{FITS Format in Planetary Surface Investigation}
\label{sec:fitspss}

\subsection{FITS Format}
FITS is an open digital standard, defined for data acquisition and archiving
in astronomical observatories in the late 70's, and it is used for
spatial telescope data.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard implemented formats in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO).
Finally, FITS format is compatible with the Planetary Data System (PDS)
archiving specifications and supported by a large number of open libraries
and software tools.

@@

\subsection{Planetary Surface Proposed Convention}
\label{ssec:geofits}
When willing to use FITS format for planetary surfaces the issue of rigorous
geographical representation must be addressed.
For their mapping and analysis needs planetary researchers commonly use
Geographical Information Systems (GIS).
In general, GIS applications excel in data interoperability even
though some have historically been anchored to Earth's spatial description.
Several of them are based on the Geospatial Data Abstraction
Library\footnote{http://www.gdal.org} (GDAL), released by the Open Source
Geospatial Foundation.
GDAL is essentially a format translation library written in C++ for geospatial raster
and vector data and, fortunately, it also offer basic support for FITS format.

@@

\begin{table}
\label{table:proj}
\caption{Projection names and codes already available in FITS suitable for
planetary applications: their correspondent in PDS4 vocabulary are
provided.}
\centering
\begin{tabular}{l l l}
\hline
FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
\hline
Zenithal Equidistant & ARC & Azimuthal Equidistant \\
Zenithal Perspective & AZP & General Vertical Near-sided Projection \\
Plate Carre & CAR & Equirectangular, Miller Cylindrical \\
Equidistant Conic & COD & Equidistant Conic \\
Conic Equal-Area & COE & Albers Conical Equal Area \\ 
Conic Ortomorphic & COO & Lambert Conformal Conic\\
Mercator & MER & Mercator, Oblique Mercator, Transverse Mercator \\
Polyconic & PCO & Polyconic \\
Sanson-Flamsteed & SFL & Sinusoidal \\
Ortographic & SIN & Orthographic \\
Stereographic & STG & Stereographic, Polar Stereographic \\
Gnomonic & TAN & Gnomonic \\
Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
\hline
\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}. We cannot find correspondences in the WCS scheme for PDS4 Robinson,} \\
\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator.}
\end{tabular}
\end{table}

@@

\subsubsection{Correspondence Between Reserved FITS Keywords and PDS4 Standard Reference}
\label{sssec:stdfitspds4}
The correspondence between mandatory and reserved FITS keywords (as defined in
the last version of the FITS standard document\footnote{https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf})
and PDS4 attributes, is described here.
\begin{enumerate}
\item{\texttt{BITPIX}.
It specifies the number of bits that represent a data value in
the associated data array.
PDS4 uses the attribute \texttt{data\_type}: correspondent 
values depend on the associated object class.}

@@

\end{enumerate}

@@

\subsubsection{Visualization of Global and Projected Maps}
High level products are meant to be used for visual inspection or GIS analysis.
They are distributed as projected rasters, often in GeoTIFF\footnote{http://geotiff.osgeo.org/}
format, as GeoTIFF make them GIS ready.
Once new keywords introduced in FITS and WCS definitions adapted to planetary surface, FITS
and GeoTIFF are completely interchangeable.
As an example we have downloaded some global maps from USGS
Astropedia\footnote{https://astrogeology.usgs.gov/search} and converted them in FITS.
We used ds9 to display the FITS image and were able to visualize coordinates in meters
as in QGIS, but also in planetary latitude and longitude thanks to the alternate coordinate
description (see figure \ref{fig:ds9qgis}, left side).
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
\caption{left: Mercury South Pole FITS image displayed in ds9;
right: GeoTIFF raster superposed to the virtual
description of the FITS raster.
Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied Physics
Laboratory, Carnegie Science}
\label{fig:ds9qgis}
\end{figure}

@@

\subsubsection{OMEGA Spectra and Geometry Data}
As stressed above, having planetary data distributed in FITS at any processing level is
a way to avoid multiple conversions in the processing chain and simplify reprocessing.
In planetary science a first spatial calibration data often consists in providing coordinates
on specific points on the detector (the center of the pixel, one or more pixel corners, etc.).
To support such look-up table coordinate representation the \texttt{TAB} algorithm
has been defined in \citet{wcsspectral}.
In the \texttt{TAB} algorithm, coordinates are listed in a coordinate array, an indexing vector
can be used to address coordinate array elements.
Coordinates can then be sampled more or less coarsely depending on the behavior of the spatial
reconstruction.
Also, when the field is only partially covered by the planetary surface, coordinate array
dimensions can be significantly reduced. 
TAB projection is implemented in the Calabretta WCSLIB\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/},
available in all major linux distribution.

As an example of non projected planetary data we have converted in FITS format data
acquired by the spectroimager OMEGA on-board of Mars Express.
OMEGA unprojected and uncalibrated data are distributed by ESA-PSA and PDS archives.
Their structure is described in the OMEGA Experiment Archive 
Interface Control Document\footnote{ftp://psa.esac.esa.int/pub/mirror/MARS-EXPRESS/OMEGA/MEX-M-OMEGA-2-EDR-FLIGHT-V1.0/DOCUMENT/EAICD\_OMEGA.PDF}.
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo4}\hspace{60pt}\includegraphics[height=17pc]{ess_marmo3}}
\caption{left: OMEGA data structure (from OMEGA EAICD); right: FITS Multi Digital Object
conversion.}
\label{fig:omegascheme}
\end{figure}

OMEGA rasters are equally sampled all over the field: we use them to exemplify
TAB description without indexing, then we only use spatial information on the center
of the pixel. 
The output result can be downloaded at the address ...

The \texttt{wcsware} tool (available in the \texttt{wcslib-utils} package in Fedora
distribution, in the \texttt{wcslib-tools} in Debian) had been run on the output file,
validating the WCS structure and checking the ability to convert to (\texttt{wcsware -w})
and from (\texttt{wcsware -x}) pixel and planetary surface coordinates.

\subsubsection{Akatsuki Imaging Data}

\subsubsection{HiRISE Mosaics}

\section{Interoperable Software Developments}
\label{sec:softdev}
VO-QGIS-plugin\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}

@@

\section{Discussion and Perspectives}
\label{sec:disc}
FITS discussion in astronomy\citep{THOMAS2015133} 
let's astronomy learn from planetary sciences too: accent must be put on interoperability
.... blabla
Planetary surface is lacking open and modular software for massive processing and can benefit
of astronomy experience. In addition to that WCS projection scheme is simpler than the
historical one used in planetary applications. 
Archiving flexibility and data model are today approached by PDS4 and EPN-TAP efforts.
Reducing custom data format conversions during the processing chain but
allowing easy conversion for final products following the needed application.
